---
title: 'Week 1 lab: Central Limit Theorem'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
output_options = c("self_contained = TRUE")
```

## Goal

Use a pseudorandom number generator to test out the CLT to see that it works (except where it doesn't)

## The plan

* Pick a distribution  
* Write a function that samples from the distribution x times and returns the sample mean  
* Run the functions many times to generate a distribution of sample means- as the sample size gets bigger, does this distribution approach a normal distribution?

Let's compute means of 2, 10, 100, and 1000 samples:

```{r}
Nvals <- c(2,10,100,1000)
```

In the code below:  
`sample_fun <- function(x) rbinom(x,1,p=0.5)`  
defines a function that returns a mean of x binomial samples. Since the binomial distribution can only return a 0 or 1, then sample(x=2) can only take values 0, 0.5, or 1.
The `for` loop runs sample_fun(x) for each value of x and plots a histogram of the sample means:

```{r, fig.width=7}
sample_fun <- function(x){
    mean(rbinom(x,1,p=0.5))
}
par(mfrow=c(2,2))
for(i in 1:length(Nvals)){
    N <- Nvals[i]
    hist(replicate(10000,sample_fun(N)), main=paste(N, 'sample means'), axes=FALSE, ylab="",xlab="")
}
```

## What's next?

Try some different distributions:

* Poisson (generally used for models of count data- you'll see it in the next lab)  
`sample <- function(x) mean(rpois(x,2))`  
* Gamma (very common as a prior distribution for Poisson and Exponential models)  
`sample <- function(x) mean(rgamma(x,7,1))`  
* Cauchy (distribution, among other things, of a ratio of Normally-distributed variables):  
`sample <- function(x) mean(rcauchy(x))`  

Notice anything different with the Cauchy? You may need to set log=True for the histogram function. What's happening here?
